On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-28 tex/draft/draft4.tex
minor overall edits

The core accretion hypothesis proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. This value, however, depends on several factors, such as the location in the disk, atmospheric opacity, and the accretion rate of solids.  Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
Our model investigates the growth of atmospheres around solid cores of fixed mass.  Without planetesimal accretion to heat the atmosphere, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction. We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk.  For different locations in the disk, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass required to nucleate atmosphere collapse is $\sim$$8.5 M_{\oplus}$ at 5 AU and steadily decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard dust opacity.   This trend is mainly explained by the cooler temperatures in the outer disk. Lower pressures and densities in the outer disk have a negligible effect on core mass requirements.  At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the atmosphere reduces the critical core mass.   The ongoing accretion of solids would delay or prevent atmospheric runaway growth.  Thus our low values for minimum core masses do not contradict the existence of ice giants, but emphasize the importance of the accretion history of solids.  We also develop a non-self-gravitating, analytic cooling model to help explain our numerical trends.  Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$20\%$ as massive as the core.

@@

This work develops quasi-static models in which the core mass is fixed and no longer accreting solids.  By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.  Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phase 2 and 3 growth in more detailed quasi-static models.  Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.

@@

This paper is organized as follows. Section \ref{sec:model} describes our model of atmospheric growth.  In Section \S\ref{sec:coolingan}, we develop a simplified analytic version of our atmospheric model.  We describe the structure and evolution of our atmosphere solutions in Section \S\ref{sec:KH}.   Section \S\ref{sec:critical} gives our final results for growth timescales and critical core masses.  We discuss some neglected effects in section \S\ref{sec:neglected} and summarize our findings in Section \S\ref{sec:conclusions}.  Some detailed derivations are presented in the appendices.

@@

The (thermodynamically isothermal) sound speed in the disk is
\begin{equation}
c\di = \sqrt{\Rg T\di} = 0.4 \sqrt{F_T} \aun{10}^{3/14} ~\text{km s}^{-1}
\end{equation}  
in terms of the specific gas constant $\Rg$.  The disk scale height is 
\begin{equation}
H\di = {c\di / \varOmega} = 0.42 \sqrt{F_T}  \, \aun{10}^{9/7} \AU\, .
\end{equation} 
in terms of the Keplerian frequency $\varOmega = \sqrt{G M_\ast/a^3}$, with $G$ the gravitational constant and $M_\ast$ the stellar (in this work Solar) mass.

We assume a dust opacity characteristic of interstellar grains, following \citet{bell94}:
\begin{equation}
\label{eq:opacitylaw}
\kappa= 2 F_\kappa  \left(\frac{T}{100\; \rm{K}}\right)^{\beta} \; \mathrm{cm^2 ~ g^{-1}},
\end{equation}
with a powerlaw index $\beta = 2$ and normalization $F_\kappa = 1$ unless noted otherwise. Grain growth tends to lower both $F_\kappa$ and $\beta$, while dust abundance scales with $F_\kappa$.   Section \S\ref{sec:op} discusses dust sublimation and more realistic opacity laws.

@@

A planet can bind a dense atmosphere if its escape velocity exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
\begin{equation}
\label{eq:RB}
\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
\end{equation}
where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius.  The scalings thus use $\mpn{10} \equiv M\pla /(10~M_\oplus)$.

Stellar tides dominate the planet's gravity beyond the Hill radius
\begin{equation}
\label{eq:RHill}
R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU ,
\end{equation}
where hydrostatic balance breaks down.  In \Eq{eq:RHill}, $M\pla$ includes mass enclosed within $\RH$.  During the early stages of evolution, $M\pla \sim M\co$, and the core mass can be used to get estimates of both $\RB$ and $\RH$.

@@

In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$ (\citealt{rafikov06}, hereafter R06).  In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.  We, however, use $\RH$ as the matching radius in both this low mass and other higher mass regimes.  This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).  Nevertheless, since the density change is modest, the choice of outer boundary condition has a similarly modest effect on our results.

@@

At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening.   See \S\ref{sec:hydro} for discussion of neglected non-hydrostatic effects on all mass scales.

Note that while $\RH$ is the outer boundary of our structure calculations, we define planet masses to include only the mass inside the smaller of $\RB$ or $\RH$.    This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.

@@

Boundary conditions must be satisfied at both the base and the top of the atmosphere with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.  In principle, our solutions describe atmospheres with $L(R\co) = 0$. In practice, since we do not directly integrate \Eq{eq:structd} we need not directly impose this boundary condition, as described in \S\ref{sec:twolayer}.

@@

To simplify our calculations of atmospheric contraction we use a two-layer model with a bottom convective region and an upper radiative layer.   The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools, the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However, the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:op}.

@@

Given the conditions at the RCB,  the density and temperature profiles along the interior adiabat,
\begin{subeqnarray}
\rho &=& \rho\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right]^{1/(\gamma -1)} \slabel{eq:rhoconv}  \\
T	&=& T\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right] \slabel{eq:Tconv}\, ,
\end{subeqnarray} 
%Note (to self) that the "1+" is neglibible in Rcb << RB limit taken elsewhere. 
follow from hydrostatic balance.  We introduce an effective Bondi radius
\begin{equation}
\RB' \equiv {G M\co \over C_PT\cb} = {\delad \over \chi} \RB
\end{equation} 
to simplify expressions, with $C_P = \Rg / \delad$ the specific heat at constant pressure.

@@

The atmosphere mass increases as radiative losses lower the internal adiabat and increase $P\cb$.  The crossover mass, $M_{\rm atm} = M\co$, is reached when
\begin{equation} \label{eq:Pcbc}
P\cb = \xi P_M ,
\end{equation} 
i.e.\ near the characteristic pressure $P_M$.  The critical value of the order unity factor $\xi$ is found by eliminating $P\cb$ from \Eqs{eq:xidef}{eq:Pcbc}.  This logarithmic factor complicates our analytic description.  Since it remains order unity, we simply hold it fixed in our scalings, but not when plotting the results.

The total energy is concentrated towards the core if $|u| \rho  r^3 \propto \rho r^2$ drops with increasing $r$.  This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ (monatomic gas) would not. 
Integration of \Eq{eq:ean} over the mass of the atmosphere thus gives
\begin{subeqnarray} 
E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{R\cb} \rho r dr \\
&\approx& - 4 \pi P\cb {\RB'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \slabel{eq:Ean} \\ 
&\approx& - 8 \pi P\cb {\RB'^{7/2} \over \sqrt{R\co}} \slabel{eq:Eus}
%&\approx& -0.31 P\cb {(\RB / \chi)^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {\RB'^{7/2} \over \sqrt{R\co}}
\end{subeqnarray} 
with $\gamma < 3/2$ and $\gamma = 7/5$ in \Eqs{eq:Ean}{eq:Eus}, respectively. % and then revert to the standard Bondi radius $\RB = G M\co/(\Rg T\di)$ in terms of the disk temperature.

@@

We present examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.  Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and validity of our two-layer cooling model is examined in \S\ref{sec:endoftime}.

@@

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop as the atmosphere becomes more massive.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.

@@

The fact that growth times have a well defined maximum is a characteristic of accelerating growth.  Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway accretion begins.  Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.  \Fig{fig:growthtime} shows that further from the star runaway growth begins at smaller $M_{\rm atm}/M\co$.  \Fig{fig:tvsM} (described in the next section) shows how the onset of runaway growth depends on core mass.

@@

The individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.  At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.  As the mass increases, the surface terms become more significant, led by the accretion energy.  However the surface terms are everywhere smaller than $L_{\rm negl}$.  Thus wherever our model is accurate -- including the crucial early phases of growth -- surface terms are a minor correction.  The neglect of surface terms in the analytic model is thus not a serious omission.

@@

\subsection{Runaway Growth Time Scale}
\label{sec:tcross}
The time to undergo atmospheric runaway growth, $t_{\rm run}$, sets a minimum timescale for the formation of giant planets.  Due to the accelerating nature of runaway growth, the precise threshold chosen for $t_{\rm run}$ (explained in \S ?) is of minor significance.

@@

Thus by restricting our attention to low masses, neglected hydrodynamic effects should be minor.   Moreover since $M_{\rm th} \propto a^{6/7}$ increases with disk radius, spherical hydrostatic models like ours have a greater range of applicability in the outer regions of disks.

@@

To derive  the global energy equation (\ref{eq:coolingglobal}) for an embedded protoplanet, we generalize the analogous calculations in stellar structure theory, e.g.\ in \S4.3 of \citet{kippenhahn90}.  For our problem, we add the effects of finite core radius, surface pressure and mass accretion.We start with the local energy equation (\ref{eq:structd}) whose more natural form in Lagrangian (mass) coordinates is $\p L/ \p m = \epsilon - T \p S /\p t$.  Integrating from the core to a higher shell with enclosed mass $M$ gives:
\begin{subeqnarray}
L - L\co &=& \int_{M\co}^M {\p L \over \p m} dm \\
&=& \int_{M\co}^M \left(\epsilon - T {\p S \over \p t} \right)dm \\
&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, ,
\end{subeqnarray} 
with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate and applying the first law of thermodynamics in the final step.

The global energy equation is derived by eliminating the partial time derivatives in \Eq{eq:DLc}, which are performed at a fixed mass,
in favor of total time derivatives, denoted with overdots.  %The physical distinction is that total derivatives include mass  accreted through the outer boundary.  
For instance, the surface radius $R$ of the shell with enclosed mass $M$ evolves as  
\begin{equation}\label{eq:Rdot}
 \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho_M},
\end{equation} 
where $\p R/\p t$ gives the Lagrangian contraction of the ``original" shell, and mass accretion through the upper boundary at rate $\dot{M}$ also changes the shell location.  
%The subscript $M$  denotes quantities at the upper boundary of total mass $M$ (though it is omitted from $M$ and $R$).  
Similarly, the volume $V = (4 \pi/3)R^3$ and pressure at the outer shell evolve as
\begin{subeqnarray}\label{eq:dot}
\dot{V}_M &=&  {\p V_{\rm M} \over \p t} + {\dot{M} \over \rho_{\rm M}}  \\
 \dot{P}_M &=& {\p P_{\rm M} \over \p t} + {\p P_M \over \p m}\dot{M} =  {\p P_{\rm M} \over \p t} - {G M  \over 4 \pi R^4} \dot{M}\, .
\end{subeqnarray} 
This derivation holds the core mass and radius fixed, $\dot{M}\co = \dot{R}\co = 0$.  Therefore the core pressure satisfies
\begin{equation}\label{eq:Pcdot}
 \dot{P}\co = \p P\co / \p t \, .
\end{equation}
The internal energy integral follows simply from  Leibniz's rule as
\begin{equation}\label{eq:udot}
\int_{M\co}^{M(t)}{\p u \over \p t} dm = \dot{U}  -  \dot{M}u_M\, .
\end{equation} 
To make further progress we use the virial theorem:
\begin{equation}
\label{eq:virial}
E_G=-3 \int_{M\co}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R\co^3 P\co),
\end{equation}
which follows from \Eqsss{eq:structb}{eq:structa}{eq:Eg} by integrating hydrostatic balance in Lagrangian coordinates.  As an aside, the integral in equation (\ref{eq:virial}) can be evaluated for a polytropic EOS to give simple expressions for the total energy:
\begin{subeqnarray}
E&=&(1-\zeta)U+4 \pi (R^3 P_M-R\co^3 P\co) \slabel{eq:vira} \\
&=&\frac{\zeta-1}{\zeta}E_G+\frac{4 \pi}{\zeta} (R^3 P_M-R\co^3 P\co) \slabel{eq:virb}
\end{subeqnarray}
where $\zeta \equiv 3(\gamma - 1)$.  We will not make this assumption and will keep the EOS general.

>>>>

The core accretion hypothesis proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. This value, however, depends on several factors, such as the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.  Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
Our model investigates the growth of atmospheres around solid cores of fixed mass.  Without planetesimal accretion to heat the atmosphere, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction. We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk.  For different locations in the disk, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass required to nucleate atmosphere collapse is $\sim$$8.5 M_{\oplus}$ at 5 AU and steadily decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard dust opacity.   This trend is mainly explained by the cooler temperatures in the outer disk. Lower pressures and densities in the outer disk have a negligible effect on core mass requirements.  At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the atmosphere reduces the critical core mass.   The ongoing accretion of solids would delay or prevent atmospheric runaway growth.  Thus our low values for minimum core masses do not contradict the existence of ice giants, but emphasize the importance of the accretion history of solids.  We also develop a non-self-gravitating, analytic cooling model to help explain our numerical trends.  Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$20\%$ as massive as the core.

@@

This work develops quasi-static models in which the core has a fixed mass and is no longer accreting solids.  By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.  Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phase 2 and 3 growth in more detailed quasi-static models.  Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.

@@

This paper is organized as follows. Section \S\ref{sec:model} describes our model of atmospheric growth.  In Section \S\ref{sec:coolingan}, we develop a simplified analytic version of our atmospheric model.  We describe the structure and evolution of our atmosphere solutions in Section \S\ref{sec:KH}.   Section \S\ref{sec:critical} gives our final results for growth timescales and critical core masses.  We discuss some neglected effects in Section \S\ref{sec:neglected} and summarize our findings in Section \S\ref{sec:conclusions}.  Some detailed derivations are presented in the appendices.

@@

The (thermodynamically isothermal) sound speed in the disk is
\begin{equation}
c\di = \sqrt{\Rg T\di} = 0.4 \sqrt{F_T} \aun{10}^{3/14} ~\text{km s}^{-1} \,,
\end{equation}  
in terms of the specific gas constant $\Rg$.  The disk scale height is 
\begin{equation}
H\di = {c\di / \varOmega} = 0.42 \sqrt{F_T}  \, \aun{10}^{9/7} \AU\, ,
\end{equation} 
in terms of the Keplerian frequency $\varOmega = \sqrt{G M_\ast/a^3}$, with $G$ the gravitational constant and $M_\ast$ the stellar (in this work Solar) mass.

We assume a dust opacity characteristic of interstellar grains, following \citet{bell94}:
\begin{equation}
\label{eq:opacitylaw}
\kappa= 2 F_\kappa  \left(\frac{T}{100\; \rm{K}}\right)^{\beta} \; \mathrm{cm^2 ~ g^{-1}},
\end{equation}
with a power-law index $\beta = 2$ and normalization $F_\kappa = 1$ unless noted otherwise. Grain growth tends to lower both $F_\kappa$ and $\beta$, while dust abundance scales with $F_\kappa$.   Section \S\ref{sec:op} discusses dust sublimation and more realistic opacity laws.

@@

A planet can bind a dense atmosphere if its escape velocity exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
\begin{equation}
\label{eq:RB}
\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU,
\end{equation}
where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius.  The scalings thus use $\mpn{10} \equiv M\pla /(10~M_\oplus)$.

Stellar tides dominate the planet's gravity beyond the Hill radius
\begin{equation}
\label{eq:RHill}
R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU ,
\end{equation}
where hydrostatic balance breaks down.  In \Eq{eq:RHill}, $M\pla$ includes the mass enclosed within $\RH$.  During the early stages of evolution, $M\pla \sim M\co$, and the core mass can be used to get estimates of both $\RB$ and $\RH$.

@@

In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$ (R06).  In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.  We, however, use $\RH$ as the matching radius in both this low mass and other higher mass regimes.  This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).  Nevertheless, since the density change is modest, the choice of outer boundary condition has a similarly modest effect on our results.

@@

At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening (see \S\ref{sec:hydro} for a discussion of neglected non-hydrostatic effects on all mass scales).

Note that while $\RH$ is the outer boundary in our structure calculations, we define planet masses to include only the mass inside the smaller of $\RB$ or $\RH$.    This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.

@@

Boundary conditions must be satisfied at both the base and the top of the atmosphere, with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.  In principle, our solutions describe atmospheres with $L(R\co) = 0$. In practice, since we do not directly integrate \Eq{eq:structd} we need not directly impose this boundary condition, as described in \S\ref{sec:twolayer}.

@@

To simplify our calculations of atmospheric contraction, we use a two-layer model with a bottom convective region and an upper radiative layer.   The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools, the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However, the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:op}.

@@

Given the conditions at the RCB,  the density and temperature profiles along the interior adiabat,
\begin{subeqnarray}
\rho &=& \rho\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right]^{1/(\gamma -1)} \slabel{eq:rhoconv}  \\
T	&=& T\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right] \slabel{eq:Tconv}\, ,
\end{subeqnarray} 
%Note (to self) that the "1+" is neglibible in Rcb << RB limit taken elsewhere. 
follow from hydrostatic balance.  We introduce an effective Bondi radius
\begin{equation}
\RB' \equiv {G M\co \over C_PT\cb} = {\delad \over \chi} \RB
\end{equation} 
to simplify expressions, with $C_P = \Rg / \delad$ the specific heat capacity at constant pressure.

@@

The atmosphere mass increases as radiative losses lower the internal adiabat and increase $P\cb$.  The crossover mass, $M_{\rm atm} = M\co$, is reached when
\begin{equation} \label{eq:Pcbc}
P\cb = \xi P_M ,
\end{equation} 
i.e.\ near the characteristic pressure $P_M$.  The critical value of the order unity factor $\xi$ is found by eliminating $P\cb$ from \Eqs{eq:xidef}{eq:Pcbc}.  This logarithmic factor complicates our analytic description.  Since it remains order unity, we simply hold it fixed in our scalings. %, but not when plotting the results.

The total energy is concentrated towards the core if $|e| \rho  r^3 \propto \rho r^2$ drops with increasing $r$.  This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ (monatomic gas) would not. 
Integration of \Eq{eq:ean} over the mass of the atmosphere thus gives
\begin{subeqnarray} 
E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{R\cb} \rho r dr \\
&\approx& - 4 \pi P\cb {\RB'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \slabel{eq:Ean} \\ 
&\approx& - 8 \pi P\cb {\RB'^{7/2} \over \sqrt{R\co}} \slabel{eq:Eus} \, ,
%&\approx& -0.31 P\cb {(\RB / \chi)^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {\RB'^{7/2} \over \sqrt{R\co}}
\end{subeqnarray} 
with $\gamma < 3/2$ and $\gamma = 7/5$ in \Eqs{eq:Ean}{eq:Eus}, respectively. % and then revert to the standard Bondi radius $\RB = G M\co/(\Rg T\di)$ in terms of the disk temperature.

@@

We present examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.  Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and the validity of our two-layer cooling model is examined in \S\ref{sec:endoftime}.

@@

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop as the atmosphere becomes more massive.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 8.99 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.

@@

The fact that growth times have a well defined maximum is a characteristic of accelerating growth.  Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway accretion starts.  Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.  \Fig{fig:growthtime} shows that further from the star runaway growth begins at smaller $M_{\rm atm}/M\co$.  \Fig{fig:tvsM} (described in the next section) shows how the onset of runaway growth depends on core mass.

@@

The individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.  At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.  As the mass increases, the surface terms become more significant, led by the accretion energy.  However, the surface terms are everywhere smaller than $L_{\rm negl}$.  Thus wherever our model is accurate -- including the crucial early phases of growth -- surface terms are a minor correction.  The neglect of surface terms in the analytic model is thus not a serious omission.

@@

\subsection{Runaway Growth Timescale}
\label{sec:tcross}
The time to undergo atmospheric runaway growth, $t_{\rm run}$, sets a minimum timescale for the formation of giant planets.  Due to the accelerating nature of runaway growth, the precise threshold chosen for $t_{\rm run}$ (explained in \S\ref{sec:timeev}) is of minor significance.

@@

Thus by restricting our attention to low masses, neglected hydrodynamic effects should be minor.   Moreover, since $M_{\rm th} \propto a^{6/7}$ increases with disk radius, spherical hydrostatic models like ours have a greater range of applicability in the outer regions of disks.

@@

To derive  the global energy equation (\ref{eq:coolingglobal}) for an embedded protoplanet, we generalize the analogous calculations in stellar structure theory, e.g.\ in \S4.3 of \citet{kippenhahn90}.  For our problem, we add the effects of finite core radius, surface pressure and mass accretion. We start with the local energy equation (\ref{eq:structd}), whose more natural form in Lagrangian (mass) coordinates is $\p L/ \p m = \epsilon - T \p S /\p t$.  Integrating from the core to a higher shell with enclosed mass $M$ gives:
\begin{subeqnarray}
L - L\co &=& \int_{M\co}^M {\p L \over \p m} dm \\
&=& \int_{M\co}^M \left(\epsilon - T {\p S \over \p t} \right)dm \\
&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, ,
\end{subeqnarray} 
with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate, and applying the first law of thermodynamics in the final step.

The global energy equation is derived by eliminating the partial time derivatives in \Eq{eq:DLc}, which are performed at a fixed mass,
in favor of total time derivatives, denoted with overdots.  %The physical distinction is that total derivatives include mass  accreted through the outer boundary.  
For instance, the surface radius $R$ of the shell with enclosed mass $M$ evolves as  
\begin{equation}\label{eq:Rdot}
 \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho_M},
\end{equation} 
where $\p R/\p t$ gives the Lagrangian contraction of the ``original" shell, and mass accretion through the upper boundary at rate $\dot{M}$ also changes the shell location.  
%The subscript $M$  denotes quantities at the upper boundary of total mass $M$ (though it is omitted from $M$ and $R$).  
Similarly, the volume $V = (4 \pi/3)R^3$ and pressure at the outer shell evolve as
\begin{subeqnarray}\label{eq:dot}
\dot{V}_M &=&  {\p V_{\rm M} \over \p t} + {\dot{M} \over \rho_{\rm M}}  \\
 \dot{P}_M &=& {\p P_{\rm M} \over \p t} + {\p P_M \over \p m}\dot{M} =  {\p P_{\rm M} \over \p t} - {G M  \over 4 \pi R^4} \dot{M}\, .
\end{subeqnarray} 
This derivation holds the core mass and radius fixed, $\dot{M}\co = \dot{R}\co = 0$.  Therefore, the core pressure satisfies
\begin{equation}\label{eq:Pcdot}
 \dot{P}\co = \p P\co / \p t \, .
\end{equation}
The internal energy integral follows simply from  Leibniz's rule as
\begin{equation}\label{eq:udot}
\int_{M\co}^{M(t)}{\p u \over \p t} dm = \dot{U}  -  \dot{M}u_M\, .
\end{equation} 
To make further progress, we use the virial theorem:
\begin{equation}
\label{eq:virial}
E_G=-3 \int_{M\co}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R\co^3 P\co),
\end{equation}
which follows from \Eqsss{eq:structb}{eq:structa}{eq:Eg} by integrating hydrostatic balance in Lagrangian coordinates.  As an aside, the integral in equation (\ref{eq:virial}) can be evaluated for a polytropic EOS to give simple expressions for the total energy:
\begin{subeqnarray}
E&=&(1-\zeta)U+4 \pi (R^3 P_M-R\co^3 P\co) \slabel{eq:vira} \\
&=&\frac{\zeta-1}{\zeta}E_G+\frac{4 \pi}{\zeta} (R^3 P_M-R\co^3 P\co) \slabel{eq:virb} \, ,
\end{subeqnarray}
where $\zeta \equiv 3(\gamma - 1)$.  We will not make this assumption and will keep the EOS general.

To express the work integral, i.e. the final term in \Eq{eq:DLc}, in terms of changes to gravitational energy, we first take the
 time derivative of \Eq{eq:virial}:
\begin{eqnarray}\label{eq:EGdot}
\dot{E}_G = 3  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho} 
 -  3{P_M \over \rho_M} \dot{M}+ 3 \dot{P}_M V_M -3 \dot{P}\co V\co  + 3  P_M {\dot{ V}_M} \, . 
\end{eqnarray} 
%where the volumes, $V_M = 4 \pi R^3/3$ and $V\co = 4 \pi R\co^3/3$. 
The first integral in \Eq{eq:EGdot} is the one we want, but the second one must be eliminated.  The time derivative of \Eq{eq:Eg} (times four) gives
\begin{subeqnarray}
 4 \dot{E}_G &=&  -4 {G M \dot{M} \over R} + 4 \int_{M\co}^M {G m \over r^2}{\p r \over \p t} dm\\ 
&=&   -4 {G M \dot{M} \over R} + 4 \pi \int_{M\co}^M r^3{\p \over \p m}{\p P \over \p t} dm \slabel{eq:4EGb} \\
&=&  -4 {G M \dot{M} \over R} -3  \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  + 3 V_M {\p P_M \over \p t} -3 V\co {\p P\co \over \p t} \slabel{eq:4EGc} \, ,
\end{subeqnarray} 
where \Eqs{eq:4EGb}{eq:4EGc} use hydrostatic balance  and integration by parts.
